---
title: "Correlating mutations with DST and lineages GLM example"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r functions, echo = F}

generate_data <- function(n, maj){
  samp <- c(0, 1)
  probs <- c(maj, 1-maj)
  probs_rev <- rev(probs)
  dst_zero <- sample(samp, n, replace = T, prob = probs)
  dst_one <- sample(samp, n, replace = T, prob = probs_rev)
  mut_zero <- sample(samp, n, replace = T, prob = probs)
  mut_one <- sample(samp, n, replace = T, prob = probs_rev)
  
  x <- data.frame(dst = c(dst_zero, dst_one), has_mut = c(mut_zero, mut_one)) 
  
  lineages <- c("lin1", "lin2", "lin3", "lin4")

  lin_zero <- sample(lineages, n, replace = T)
  lin_one <- sample(lineages, n, replace = T, prob = c(maj, rep((1-maj)/3, 3)))
  
  x$lin <- c(lin_zero, lin_one)
  
  x
}


```

### Introduction


### Set up the data and parameters

n samples where the majority are DST -ve (0)
n samples where the majority are DST +ve (1)
n samples where the majority do not have a mutation (0)
n samples where the majority have a mutation (1)

```{r params}

n <- 100
maj <- 0.9
samp <- c(0, 1)
probs <- c(maj, 1-maj)
probs_rev <- rev(probs)
dst_zero <- sample(samp, n, replace = T, prob = probs)
dst_one <- sample(samp, n, replace = T, prob = probs_rev)
mut_zero <- sample(samp, n, replace = T, prob = probs)
mut_one <- sample(samp, n, replace = T, prob = probs_rev)

x <- data.frame(dst = c(dst_zero, dst_one), has_mut = c(mut_zero, mut_one)) 

head(x, 10)
tail(x, 10)

cor(x)
```

### Run a model on this data 

```{r first-model}
model <- glm(formula = dst ~ has_mut, data = x, family = binomial(link = "logit"))

summary(model)

cols <- ifelse(x$dst == 1, "red", "blue")
jit <- 0.5
plot(jitter(x$dst, jit), jitter(x$has_mut, jit), col = cols)
x_dst <- seq(0, 1, 0.01)
y_has_mut <- predict(model, list(has_mut = x_dst),type="response")
lines(x_dst, y_has_mut)

```

DST and presence of mutation are highly correlated. Model result is highly sig. 

### Add lineages as co-variates

Make lineage 1 highly correlated with presence of mutation (1) and DST positive (1)

```{r add-lineages}

lineages <- c("lin1", "lin2", "lin3", "lin4")

lin_zero <- sample(lineages, n, replace = T)
lin_one <- sample(lineages, n, replace = T, prob = c(maj, rep((1-maj)/3, 3)))

x$lin <- c(lin_zero, lin_one)

head(x, 10)
tail(x, 10)

```

First run a model predicting DST from lineage

```{r dst-lin}

model_2 <- glm(formula = dst ~ lin - 1, data = x, family = binomial(link = "logit"))

summary(model_2)

```

Model predicting DST from lineage and mutation

```{r dst-lin-plus-mut}

model_3 <- glm(formula = dst ~ lin + has_mut - 1, data = x,  family = binomial(link = "logit"))

summary(model_3)

```

Is model_3 significantly different from the first model?

```{r model-sig-diff}

anova(model_3, model, test='Chisq')

```

### Problems/questions

1. why are the results so different every time?

Something like 10-20% of the models come out as non-signifcant when the lineages are added (DST ~ has mutation + lineages, compared to DST ~ has mutation only)

```{r models-loop}

models_list <- list()
loops <- 100
for (i in 1:loops){
  data <- generate_data(n, maj)
  model_loop <- glm(formula = dst ~ lin + has_mut - 1, data = data,  family = binomial(link = "logit"))
  has_mut_coef <- summary(model_loop)$coefficients["has_mut", "Estimate"]
  has_mut_p <- summary(model_loop)$coefficients["has_mut", "Pr(>|z|)"]
  anova_test <- anova(model_loop, model, test='Chisq')  
  anova_test_p <- anova_test$`Pr(>Chi)`[2]
  models_list[[i]] <- round(c(has_mut_coef, has_mut_p, anova_test_p), 3)
}

model_df <- setNames(data.frame(do.call("rbind", models_list)), c("has_mut_coef", "has_mut_p", "model_p"))
model_df$non_sig <- ifelse(model_df$model_p > 0.05, 1, 0)
model_df_non_sig <- subset(model_df, non_sig == 1)
model_df_non_sig
nrow(model_df_non_sig)/nrow(model_df)

```

2. what are the relationships between lineage 1 coefficient/p-value and the mutation coefficient/p-value?

The p-value for lineage 1 seems to vary a lot and the p-value for the presence of a mutation is always significant.

Shouldn't the p-value for the mutation become non-significant in the presence of a highly-correlated variable (lineage)?

```{r models-loop-2}

models_list <- list()
loops <- 10
for (i in 1:loops){
  data <- generate_data(n, maj)
  model_loop <- glm(formula = dst ~ lin + has_mut - 1, data = data,  family = binomial(link = "logit"))
  lin_coef <- summary(model_loop)$coefficients["linlin1", "Estimate"]
  lin_p <- summary(model_loop)$coefficients["linlin1", "Pr(>|z|)"]
  has_mut_coef <- summary(model_loop)$coefficients["has_mut", "Estimate"]
  has_mut_p <- summary(model_loop)$coefficients["has_mut", "Pr(>|z|)"]
  models_list[[i]] <- round(c(lin_coef, lin_p, has_mut_coef, has_mut_p), 3)
}

model_df <- setNames(data.frame(do.call("rbind", models_list)), c("lin_1_coef", "lin_1_p", "has_mut_coef", "has_mut_p"))
model_df

```












